Expos\'{e} Master Thesis\\ Checking Assertions for Smart Contracts

### 2020-11-16 meeting-notes/20200928-expose.tex
revised entrypoints, transforming assertions

\section{Suggestions (Tezos)}
\label{sec:suggestions-tezos}

The syntax for assertions should be chosen to be similar to OCaml.
A proposal for the assertion in \eqref{eq:5} could be
\begin{lstlisting}[language=caml]
parameter (_, (_, p)) =
  forall (n : int)
    if (2 <= n && n <= p / 2)
      p % n <> 0
\end{lstlisting}
Such a parameter phrase may occur multiple times. Its single argument
is a pattern that matches the contract's argument. The type of the
pattern variables (here just \lstinline/p/) is determined from the code of the
contract.

Multiple parameter phrases can be used to state assertion for
contracts with multiple entry points. In Tezos/Michelson, entry points are
modeled by having a sum type as parameter type. For example, a
contract with two entry points can be equipped with assertions as
follows.
\begin{lstlisting}[language=caml]
parameter (Left x) =
  0 < x && x < 100

@@

>>>>

\section{Suggestions}

\subsection{General}
\label{sec:suggestions-general}

In the example, we checked if all elements satisfy a certain condition:
\begin{lstlisting}[language=Caml]
forall (n : Int)
   if ( 2 <= n && n <= sqrt(p))
       assert(p%n != 0)
\end{lstlisting}
However, if we consider  parameter p to be a sorted list and we want
to assert that it is indeed sorted, we would need to check two values: 
\begin{lstlisting}[language=Caml]
forall (n : Int, m: Int)
   if ( n > m)
       assert(p.(n) > p.(m))
\end{lstlisting}

You can have foralls with more than one variable, but you could also nest the foralls as in

\begin{lstlisting}[language=Caml]
forall( n : int)
 forall(m : int) ...
\end{lstlisting}
Externally, I think your proposed syntax is nicer, but internally it will be simpler to just have 
\lstinline/Forall( ident, type, expr)/
in the type for the AST.

Somehow the smart random generator will have to know in which range to
generate the values.  Considering the second example, the generator should generate 
\begin{lstlisting}[language=Caml]
m = random(0, length(list))
n = random(m, length(list))
\end{lstlisting}
For this transformation, the assertion should be completely explicit.
That is, in the example, there should be an additional condition

\begin{lstlisting}[language=Caml]
... if (n < length(p) && m < length(p)) ...
\end{lstlisting}
which means we should have the length of a list available as a primitive.

Can we realize this transformation from the above syntax, or do we
need additional syntax that expresses this transformation separately? 
Do I even have to consider these things when creating the grammar?

The process could be something like this:
\begin{enumerate}
\item  calculate the negation of the formula. This amounts to
  swapping forall and exists, skipping the conditions, and negating
  the body of the assertion. This last step should also be done by
  applying de Morgan exhaustively down to the literals. In your
  example you'd end up with

\begin{lstlisting}[language=Caml]
exists(n : int)
 exists(m : int)
    if (n< length(p) && m < length(p))
      if (n>m)
         check( p.(n) <= p.(m) )
\end{lstlisting}
\item look at the atomic constraints that bound the exists and move them upfront. Example: 
  \begin{itemize}
  \item  the first constraint is \lstinline/n<length(p)/ $\to$ you need to break up
    the conjunction
  \item independent of \lstinline/exists(m)/, hence move backward
  \end{itemize}
\begin{lstlisting}[language=Caml]
exists(n : int)
 if (n < length(p))
   exists(m : int)
     if (m < length(p))
       ...
\end{lstlisting}
\item  if the constraint meets its generator (exists), then merge it:

\begin{lstlisting}[language=Caml]
exists(n : int, n < length(p))
\end{lstlisting}
\item  same for m

\begin{lstlisting}[language=Caml]
 exists(m : int, m < length(p))
\end{lstlisting}
( at this point, each generator / exists could come with a number of upper and lower bounds )
\item now we need to consider this

\begin{lstlisting}[language=Caml]
exists(n : int, n < length(p))
 exists(m : int, m < length(p))
   if (n>m)
     ...
\end{lstlisting}
this condition gives us an upper bound on m (this requires some manipulation of the formula, in general)
so we just add it as another bound to m:

\begin{lstlisting}[language=Caml]
exists(n : int, n < length(p))
 exists(m : int, m < length(p) && m < n)
    ...
\end{lstlisting}
\item  you can find out (but this is more involved and thus optional)
  that \lstinline/0 < n/ is required for the \lstinline/exists(m ...)/ to be nonempty and that
  \lstinline/m < n/ implies \lstinline/m < length(p)/:

\begin{lstlisting}[language=Caml]
exists(n : int, 0 < n,  n < length(p))
 exists(m : int, 0 <=m, m < n)
   check( p.(n) <= p.(m) )
\end{lstlisting}
At this point, you can read off the invocation for the random generator.
If any constraints remain, they'd have to be checked at run time.
In the best case, things work out like in the example. Otherwise, each if-constaint that turns out to be false results in a useless test run.
\end{enumerate}

For background, there is a nice paper about the underlying testing methodology:
How to specify it! by John Hughes:
\url{https://www.tfp2019.org/resources/tfp2019-how-to-specify-it.pdf}

\subsection{Tezos}
\label{sec:suggestions-tezos}

Michelson and the Tezos infrastructure directly support entry points
with a specific notation. Hence, the syntax should reflect and reuse
that notation.

The syntax for assertions should be chosen to be similar to OCaml.
A proposal for the assertion in \eqref{eq:5} could be
\begin{lstlisting}[language=caml]
entrypoint %default
  parameter (_, (_, p)) =
    forall (n : int)
      if (2 <= n && n <= p / 2)
        p % n <> 0
\end{lstlisting}
A parameter phrase may occur multiple times. Its single argument
is a pattern that matches the contract's entrypoint's argument. The type of the
pattern variables (here just \lstinline/p/) is determined from the code of the
contract.

The entrypoint phrase may also occur multiple times. However, it is
illegal to have patterns that overlap, i.e., which match the same
contract invocation. Overlapping may occur due to a repeated pattern
in clauses for the same entrypoint, but also due to a pattern for the
\lstinline/%default/ overlapping with a pattern for another entrypoint.

In Tezos/Michelson, entry points are
modeled by having a sum type as parameter type. For example, a
contract with two entry points can be equipped with assertions as
follows.
\begin{lstlisting}[language=caml]
parameter (Left x) =
  0 < x && x < 100

@@

These phrases would refer to the \lstinline/%default/ entrypoint, as
none is explicitly specified. Suppose the parameter type for the
contract in question was specified in Michelson as follows.
\begin{lstlisting}
parameter (or (int %set_scale) (pair int int %set_triangle))
\end{lstlisting}
An equivalent way to stating the previous assertions would be to use
entrypoints.
\begin{lstlisting}[language=caml]
entrypoint %set_scale
  parameter (x) =
    0 < x && x < 100

entrypoint %set_triangle
  parameter (a, b) =
    exists (c : int)
      if (0 < c && c <= a + b)
        a*a + b*b = c*c
\end{lstlisting}
